---
title: "2.exploration"
author: "Joao Lopes"
date: "2024-04-18"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)

```

\newpage

\tableofcontents

***
\newpage

# 1. VARIATION

[from https://r4ds.hadley.nz/eda#variation]

```{r 1_variation, warning=FALSE, message=FALSE}
library("tidyverse")
library("nycflights13")

```

## 1.1. TYPICAL VALUES

```{r 1_1_variation}
#visualize distribution
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.5)

#visualize distribution in detail
diamonds |> 
  filter(carat < 3) |>                 #filter in smaller diamonds
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(xintercept=c(seq(0,3,0.5)),linetype = "dashed",color="red")

#visualize distribution
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5)
  
#visualize distribution in detail
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))

```

## 1.2. UNUSUAL VALUES

```{r 1_2_variation}
#look at the data
diamonds |> 
  filter(between(y, 3, 20)) |>         #filter out unusual values
  select(price, x, y, z) |>
  arrange(y)

#drop entire row of data
diamonds2 <- diamonds |> 
  filter(between(y, 3, 20))

#replacing with missing values
diamonds2 <- diamonds |> 
  mutate(y = if_else(y < 3 | y > 20, NA, y))

ggplot(diamonds2, aes(x = x, y = y)) + 
  geom_point()

ggplot(diamonds2, aes(x = x, y = y)) + 
  geom_point(na.rm = TRUE)

#meaningful missing values
flights |> 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + (sched_min / 60)
  ) |> 
  ggplot(aes(x = sched_dep_time)) + 
  geom_density(aes(color = cancelled), bw = 1/4)

```

***
\newpage

# 2. MISSING VALUES

[from https://r4ds.hadley.nz/missing-values]

```{r 2_missing, warning=FALSE, message=FALSE}
library("tidyverse")
library("nycflights13")

```

## 2.1. EXPLICIT MISSING VALUES

```{r 2_1_1_missing}
treatment <- tribble(
  ~person,           ~treatment, ~response,
  "Derrick Whitmore", 1,         7,
  NA,                 2,         10,
  NA,                 3,         NA,
  "Katherine Burke",  1,         4
)

```

### Last observation carried forward

```{r 2_1_2_missing}
treatment |>
  fill(everything())

```

### Fixed values

```{r 2_1_3_missing}
treatment |>
  mutate(response = coalesce(response, 0)) |>
  fill(everything())

```

### Numeric value represents NA

```{r 2_1_4_missing}
csv <- "
person,          treatment, response
Derrick Whitmore,        1,        7
Derrick Whitmore,        2,       10
Derrick Whitmore,        3,       99
Katherine Burke,         1,        4"
read_csv(csv, na = "99")

read_csv(csv) |>
  mutate(response = na_if(response, 99))

```

### NaN

```{r 2_1_5_missing}
v <- c(NA, NaN)
v * 10
v == 1
is.na(v)
is.nan(v)

0 / 0 
0 * Inf
Inf - Inf
sqrt(-1)

```

## 2.2. IMPLICIT MISSING VALUES

```{r 2_2_missing}
stocks <- tibble(
  year  = c(2020, 2020, 2020, 2020, 2021, 2021, 2021),
  qtr   = c(   1,    2,    3,    4,    2,    3,    4),
  price = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
#1. Price in 2020 Q4 is explicitly missing
#2. Price in 2021 Q1 is implicitly missing

```

### Pivoting

```{r 2_2_1_missing}
stocks |>
  pivot_wider(
    names_from = qtr, 
    values_from = price
  ) |>
  pivot_longer(
    cols = -year,
    names_to = "qtr",
    values_to = "price"
  )

```

### Complete

```{r 2_2_2_missing}
stocks |>
  complete(year, qtr)

stocks |>
  complete(year = 2019:2021, qtr)

```

### Joins

```{r 2_2_3_missing}
flights |> 
  distinct(faa = dest) |>              #calculate the distinct "dest"
  anti_join(airports)                  #obtain absent "dest"

flights |> 
  distinct(tailnum) |>                 #calculate the distinct "dest"
  anti_join(planes)                    #obtain absent "dest"

```

## 2.3. FACTORS AND EMPTY GROUPS

```{r 2_3_missing}
health <- tibble(
  name   = c("Ikaia", "Oletta", "Leriah", "Dashay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age    = c(34, 88, 75, 47, 56),
)

#empty groups with count
health |> count(smoker)
health |> count(smoker, .drop = FALSE)

#empty groups with ggplot
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete()
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)

#empty groups with group_by()
health |> 
  group_by(smoker, .drop = FALSE) |> 
  summarize(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  )
health |> 
  group_by(smoker) |> 
  summarize(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age),
    sd_age = sd(age)
  ) |> 
  complete(smoker)
 
```

***
\newpage
  
# 3. COVARIATION

[from https://r4ds.hadley.nz/eda#covariation]

```{r 3_covariation, warning=FALSE, message=FALSE}
library("tidyverse")
library("hexbin")

```

## 3.1. CATEGORICAL AND NUMERICAL VARIABLES

```{r 3_1_covariation}
#use geom_freqpoly()
ggplot(diamonds, aes(x = price, color = cut)) + 
  geom_freqpoly(binwidth = 500, linewidth = 0.75)

#use geom_freqpoly() with densities
ggplot(diamonds, aes(x = price, y = after_stat(density), color = cut)) + 
  geom_freqpoly(binwidth = 500, linewidth = 0.75)  

#use geom_density()
ggplot(diamonds, aes(x = price, color = cut)) + 
  geom_density(bw = 500, linewidth = 0.75)  

#use geom_boxplot()
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot()
 
```

## 3.2. TWO CATEGORICAL VARIABLES

```{r 3_2_covariation}
#geom_count()
ggplot(diamonds, aes(x = cut, y = color)) +
  geom_count()

#geom_tile()  
diamonds |> 
  count(color, cut) |>
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))
 
```

## 3.3. TWO NUMERICAL VARIABLES

```{r 3_3_covariation}

smaller <- diamonds |> 
  filter(carat < 3)                    #filter in smaller diamonds

#geom_point()
ggplot(smaller, aes(x = carat, y = price)) +
  geom_point()

#geom_point() with alpha
ggplot(smaller, aes(x = carat, y = price)) + 
  geom_point(alpha = 0.01)

#geom_bin2d()
ggplot(smaller, aes(x = carat, y = price)) +
  geom_bin2d()

#geom_hex()
ggplot(smaller, aes(x = carat, y = price)) +
  geom_hex()

#geom_boxplot()
ggplot(smaller, aes(x = carat, y = price)) + 
  geom_boxplot(aes(group = cut_width(carat, 0.1)))

```

***
\newpage
  
# 4. PATTERNS AND MODELS

[from https://r4ds.hadley.nz/eda#patterns-and-models]

```{r 4_patterns, warning=FALSE, message=FALSE}
library("tidyverse")
library("tidymodels")

```

```{r 4_1_patterns}
diamonds <- diamonds |>
  mutate(
    log_price = log(price),
    log_carat = log(carat)
  )

#functions linear_reg() and fit()
diamonds_fit <- linear_reg() |>
  fit(log_price ~ log_carat, data = diamonds)

#function augment()
diamonds_aug <- augment(diamonds_fit, new_data = diamonds) |>
  mutate(.resid = exp(.resid))

#unexplained variation in price using carat
ggplot(diamonds_aug, aes(x = carat, y = .resid)) + 
  geom_point()

#modelling unexplained variation in price using cut
ggplot(diamonds_aug, aes(x = cut, y = .resid)) + 
  geom_boxplot()

```
